Finances and Individual Well Being

### 2023-03-17 outputs/paper/paper3.qmd
final copy

To first get a better grasp of this investigation, we conducted a small literature review on the relationships between finances and well being in the United States. A common well known relationship between well being and finances is the positive correlation between Life Satisfaction and Economic well being [@helliwell2012world]. Furthermore, there are other economic theories that also attempt to interrogate the relationship between happiness and economic well being such as the Easterlin Paradox [@easterlin1974does].
In this regard, the positive correlation between finances and happiness is well understood. However, looking deeper into the literature, we can also identify how the deprivation of economic well being can also negatively impact an individual’s quality of life. Specifically, research shows that economic hardship can impact marital quality and cause instability. This is due to an increase in occupational and financial stress that generate frustration, stress, anger and depression. [@conger1990linking] Furthermore, economic hardships can also inflict long-term consequences on children psychological well being in adulthood. [@sobolewski2005economic]. Finally, economic hardships play a major role in increasing depression levels, as depression can emerge as a byproduct of low income, education, being young, and having young children. [@ross1985hardship]
	Based on our brief literature review and the current economic climate of 2023, we were inspired to engage with the GSS 2021 survey data to answer questions about the impact of financial satisfaction with overall well being.

# 2 Data
This report focuses on the General Social Survey (GSS), which is a survey that observes and monitors public opinion and behavior in the United States ("US General Social Survey", 2021). Since 1972, the GSS has been conducted by the NORC at the University of Chicago, and is currently funded by the National Science Foundation (NSF) This survey was conducted annually until 1990, where it shifted towards a bi-annual schedule. Its aim is to maintain a consistent set of questions that aggregates American sentiments through similar sampling and questioning approaches. Depending on the year, political context, and contemporary academic practices, additional questions may be added or altered. However, there are still a core set of questions that are maintained and given every year. These core survey questions are mostly identity related (age, income, gender, etc.), but a few are also relevant to general opinions and well being. Additionally, the survey consists of repeating and topical modules, with each participant receiving a subset of both. For the scope of this study, we are going to look at survey responses produced from the 2021 survey.

@@

## 2.2 Survey Biases and Ethics
The lack of an interviewer implied the possibility of non-response, as participants are less socially pressured to fill out responses they may or may not feel inclined to report. To address this, an introductory form was administered to all participating adults which somewhat ensured that those taking the survey are dedicated enough to offer adequate responses.This was a trade off in the sampling approach that must be noted compared to previous years. Therefore, weights were carefully placed to account for factors such as population totals, density of surveyed addresses nearby, and non-response. The weighting system must account for new data and be similar to previous years to avoid skewing the results. However, despite all of these measures and countermeasures, the greatest challenge against the GSS (in all years) is the threat of survey bias, as all the data collected for all surveys are from participants willing to fill out a 90 minute survey. This could mean that the only people represented in the GSS survey are those with the free time and willingness to offer survey responses to NORC, potentially misrepresenting the overall population.

@@

We also currently have variable "- LIFE: "Subjective opinion of respondent's view on life as dull, routine, or exciting"  included in the data extract. We ended up not working with this variable as it only included samples from two out of the 3 ballots employed in 2021.

@@

According to @fig-satfincome, when comparing the financial satisfaction of respondents with their income ranges, the number of respondents from the $0 - $89,999 ranges that claim they are not satisfied at all are relatively similar. However, when examining the total number of respondents within their respective income ranges, it can be seen that it increases as the income range increases. This indicates that the proportion of respondents that do not feel satisfied at all decreases as the income range increases. Conversely, more respondents claim that they are either more or less satisfied or pretty well satisfied as the income range increases, and are heavily skewed to those that have higher income ranges.

@@

The comparison between the respondents’ happiness and financial satisfaction are visualized on @fig-satfinhappy. We observe that the number of respondents that claimed that they are generally happy increased as the financial satisfaction of these respondents increased as well. A higher proportion of respondents that claim they are not too happy also respond that they are not financially satisfied, and so there seems to be a pattern of general happiness being linked to financial satisfaction.

@@

The observation of respondents that describe their level of happiness compared to their income is demonstrated on @fig-happyincome. Each income group shows relatively similar proportions of respondents that claim that they are pretty happy, with fewer proportions claiming they are not too happy or very happy. However, as the income range increases, more respondents claim that they are pretty happy or very happy. There is an interesting observation with those families with a combined income of $40,000 to $89,999. This income group seems to have a slightly disproportionate amount of respondents saying they are not too happy when comparing with the proportions of other income groups.

@@

From observing @fig-hapmarincome, the amount of respondents that are not too happy with their marriage are in the income range of 40,000 to 90,000. The number of not too happy responses is seen to increase as does income, however it falls slightly in the income range of $90,000 to $170,000 from the previous group. Additionally, very happy marriages are seen to be steadily increasing along with the income.

@@

From observing @fig-healthincome as income increases so does the number of excellent health responses. The three income ranges that fit into the category of below $40,000 all have a relatively low amount of excellent health responses, and a moderate amount of good and fair health. However, from the income ranges that include $40,000 and more, the excellent health response is much larger, meanwhile good health occupies the majority of responses. Additionally, as income increases the poor health responses are also seen to slightly decrease.

@@

While looking at the @fig-finaltsatfin, we have observed respondents that are pretty well satisfied with their financial situation that claim that their changes in wealth have gotten better, and almost none that are satisfied claim that their wealth has gotten worse. Additionally, more than half of those not satisfied with their financial situation claim that their financial situations have gotten worse over time, and only less than a minority are not satisfied at all as their changes in wealth have gotten better. The majority of the respondents that claim that they are more or less satisfied with their finances claim that their changes in wealth have either gotten better or stayed the same, and only a minority believes that they have gotten worse. Ultimately those that have either experienced a better or the same change in wealth are more likely to be either more or less satisfied or pretty well satisfied. Meanwhile, whore who have experienced a worse change of wealth have become less satisfied.  
\newpage

@@

According to @fig-finatincome, the majority of respondents who claim that their changes in wealth have gotten better, fall into the $40,000 to $89,999 as well as the $90,000 to $170,000 income ranges. Additionally, a good portion of the respondents in the following two income ranges, claim that their wealth has remained the same, and only a minority believes that their wealth has gotten worse. On the other end of the graph, the income groups ranging from $0 to $9,999, $10,000 to $19,999 and $20,000 to $39,000 are all almost fairly equally split between believing that their wealth changes are either better, the same, or worse.

@@

The relationship between financial satisfaction and changes in income is very interesting. While changes in wealth and income indicated an even proportion of individuals that had incomes that improved, stayed the same, or gotten better, respondents that experienced worse changes in wealth over-represent those who are not satisfied with their financial situations. This bias is interesting, as it demonstrates the sentiment-based nature of the variable of financial satisfaction. In this regard, a more accurate survey to probe financial well-being should be targeted towards one’s financial ability to make ends meet.

@@

>>>>

To first get a better grasp of this investigation, we conducted a small literature review on the relationships between finances and well being in the United States. A common well known relationship between well being and finances is the positive correlation between Life Satisfaction and Economic well being [@helliwell2012world]. Furthermore, there are other economic theories that also attempt to interrogate the relationship between happiness and economic well being such as the Easterlin Paradox [@easterlin1974does]. In this regard, the positive correlation between finances and happiness is well understood. However, looking deeper into the literature, we can also identify how the deprivation of economic well being can also negatively impact an individual's quality of life. Specifically, research shows that economic hardship can impact marital quality and cause instability. This is due to an increase in occupational and financial stress that generate frustration, stress, anger and depression. [@conger1990linking] Furthermore, economic hardships can also inflict long-term consequences on children psychological well being in adulthood. [@sobolewski2005economic]. Finally, economic hardships play a major role in increasing depression levels, as depression can emerge as a byproduct of low income, education, being young, and having young children. [@ross1985hardship] Based on our brief literature review and the current economic climate of 2023, we were inspired to engage with the GSS 2021 survey data to answer questions about the impact of financial satisfaction with overall well being.

# 2 Data

@@

## 2.2 Survey Biases and Ethics

The lack of an interviewer implied the possibility of non-response, as participants are less socially pressured to fill out responses they may or may not feel inclined to report. To address this, an introductory form was administered to all participating adults which somewhat ensured that those taking the survey are dedicated enough to offer adequate responses.This was a trade off in the sampling approach that must be noted compared to previous years. Therefore, weights were carefully placed to account for factors such as population totals, density of surveyed addresses nearby, and non-response. The weighting system must account for new data and be similar to previous years to avoid skewing the results. However, despite all of these measures and countermeasures, the greatest challenge against the GSS (in all years) is the threat of survey bias, as all the data collected for all surveys are from participants willing to fill out a 90 minute survey. This could mean that the only people represented in the GSS survey are those with the free time and willingness to offer survey responses to NORC, potentially misrepresenting the overall population.

@@

We also currently have variable "- LIFE:"Subjective opinion of respondent's view on life as dull, routine, or exciting" included in the data extract. We ended up not working with this variable as it only included samples from two out of the 3 ballots employed in 2021.

@@

According to @fig-satfincome, when comparing the financial satisfaction of respondents with their income ranges, the number of respondents from the \$0 - \$89,999 ranges that claim they are not satisfied at all are relatively similar. However, when examining the total number of respondents within their respective income ranges, it can be seen that it increases as the income range increases. This indicates that the proportion of respondents that do not feel satisfied at all decreases as the income range increases. Conversely, more respondents claim that they are either more or less satisfied or pretty well satisfied as the income range increases, and are heavily skewed to those that have higher income ranges.

@@

The comparison between the respondents' happiness and financial satisfaction are visualized on @fig-satfinhappy. We observe that the number of respondents that claimed that they are generally happy increased as the financial satisfaction of these respondents increased as well. A higher proportion of respondents that claim they are not too happy also respond that they are not financially satisfied, and so there seems to be a pattern of general happiness being linked to financial satisfaction.

@@

The observation of respondents that describe their level of happiness compared to their income is demonstrated on @fig-happyincome. Each income group shows relatively similar proportions of respondents that claim that they are pretty happy, with fewer proportions claiming they are not too happy or very happy. However, as the income range increases, more respondents claim that they are pretty happy or very happy. There is an interesting observation with those families with a combined income of \$40,000 to \$89,999. This income group seems to have a slightly disproportionate amount of respondents saying they are not too happy when comparing with the proportions of other income groups.

@@

From observing @fig-hapmarincome, the amount of respondents that are not too happy with their marriage are in the income range of 40,000 to 90,000. The number of not too happy responses is seen to increase as does income, however it falls slightly in the income range of \$90,000 to \$170,000 from the previous group. Additionally, very happy marriages are seen to be steadily increasing along with the income.

@@

From observing @fig-healthincome as income increases so does the number of excellent health responses. The three income ranges that fit into the category of below \$40,000 all have a relatively low amount of excellent health responses, and a moderate amount of good and fair health. However, from the income ranges that include \$40,000 and more, the excellent health response is much larger, meanwhile good health occupies the majority of responses. Additionally, as income increases the poor health responses are also seen to slightly decrease.

@@

While looking at the @fig-finaltsatfin, we have observed respondents that are pretty well satisfied with their financial situation that claim that their changes in wealth have gotten better, and almost none that are satisfied claim that their wealth has gotten worse. Additionally, more than half of those not satisfied with their financial situation claim that their financial situations have gotten worse over time, and only less than a minority are not satisfied at all as their changes in wealth have gotten better. The majority of the respondents that claim that they are more or less satisfied with their finances claim that their changes in wealth have either gotten better or stayed the same, and only a minority believes that they have gotten worse. Ultimately those that have either experienced a better or the same change in wealth are more likely to be either more or less satisfied or pretty well satisfied. Meanwhile, whore who have experienced a worse change of wealth have become less satisfied.\
\newpage

@@

According to @fig-finatincome, the majority of respondents who claim that their changes in wealth have gotten better, fall into the \$40,000 to \$89,999 as well as the \$90,000 to \$170,000 income ranges. Additionally, a good portion of the respondents in the following two income ranges, claim that their wealth has remained the same, and only a minority believes that their wealth has gotten worse. On the other end of the graph, the income groups ranging from \$0 to \$9,999, \$10,000 to \$19,999 and \$20,000 to \$39,000 are all almost fairly equally split between believing that their wealth changes are either better, the same, or worse.

@@

The relationship between financial satisfaction and changes in income is very interesting. While changes in wealth and income indicated an even proportion of individuals that had incomes that improved, stayed the same, or gotten better, respondents that experienced worse changes in wealth over-represent those who are not satisfied with their financial situations. This bias is interesting, as it demonstrates the sentiment-based nature of the variable of financial satisfaction. In this regard, a more accurate survey to probe financial well-being should be targeted towards one's financial ability to make ends meet.

@@

## 5.1 Survey Link

https://docs.google.com/forms/d/e/1FAIpQLSc9RScig9mAIUceeQA5w4LVJeSN8PfxJNavqAVnnWPrTHeejg/viewform

## 5.2 Survey Questions

![](images/paste-C08F7830.png)

![](images/paste-F617C5EE.png)

![](images/paste-314E528F.png)

![](images/paste-DF7947C0.png)

![![](images/paste-9EB2CCCB.png)](images/paste-0B2CB8F3.png)
